We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-02-08 paper/paper.tex
references

\begin{abstract}
We use numerical simulations to build an explicit probability
distribution for the asphericity in the satellite distribution around
galaxies similar to the Local Group (LG) in the Lambda Cold Dark
Matter (LCDM) paradigm. 
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-Dark and ELVIS. 
Detailed results differ greatly among the simulations, suggesting a
strong influence of the typical DM halo mass in the different pair samples.
However, there are three common trends in these simulations. 
First, at most of $2\%$ the pairs are expected to have satellite
distributions with the same asphericity as the Local Group; second,
between $27\%$ to $56\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $3\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in M31 galaxy turn out to have a rather
typical distribution in the LCDM context, while the MW is atypical in
this respect.  
These results place the LG at the level of a $3\sigma$ outlier
in the LCDM paradigm. 
The approach presented here facilitates the comparison between
asphericity results from different numerical setups and choices to
select halo pairs in simulations, hopefully helping out to clarify the
reasons behind the LG atipicallity in LCDM. 
\end{abstract}

@@

The spatial distribution of satellite galaxies around our Milky Way and the
M31 galaxy provides a stringent test for structure formation theories
in a explicit cosmological context. 
The presence of a Vast Polar Structure of satellites around the Milky
Way has been stablished in the last decade.
This structure is present in the $11$ brightest "classical" dwarf
galaxies and has been confirmed as fainter satellites have been
detected. 
In the case of M31,  observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of an homogeneous
sample of $27$ satellites. 
However, taking into account the $11$ brightest satellites (as a
comparison with the MW) there does not seem to be any planar structure.

The question whether this kind of planar configurations are atypical
in the observable Universe cannot be properly addressed given the
current limits in observations. 
So far it has only been possible to contraint that $5\%$ of the
Milky-Way like galaxies in the Universe have two bright satellite
galaxies as is the case of the Milky Way and its Large and Small
Magellanic Clouds.

>>>>

\begin{abstract}
We use numerical simulations to build an explicit probability
distribution for the asphericity in the satellite distribution around
galaxies similar to the Local Group (LG) in the Lambda Cold Dark
Matter (LCDM) paradigm. 
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-Dark and ELVIS. 
Detailed results differ among the simulations suggesting a strong
influence of the typical DM halo mass in the different pair samples.  
However, there are three common trends in these simulations. 
First, at most of $2\%$ the pairs are expected to have satellite
distributions with the same asphericity as the LG; second,
between $27\%$ to $56\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $3\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in M31 galaxy turn out to have a rather
typical distribution in the LCDM context, while the MW is atypical in
this respect.  
These results place the LG at the level of a $3\sigma$ outlier
in the LCDM paradigm. 
The approach presented here facilitates the comparison between
different numerical setups and choices to select halo pairs in simulations.
Quantifying the probability density for the satellite distribuition
asphericity as a function of halo mass and large scale environment 
will be a necessary step to understand the origin of the LG
atipicallity in the LCDM context as presented here.  
\end{abstract}

@@

The spatial distribution of satellite galaxies around our Milky Way
(MW) and the M31 galaxy provides a stringent test for structure
formation theories in a explicit cosmological context.  
The presence of a Vast Polar Structure of satellites around the Milky
Way has been stablished in the last decade.
This structure is present in the $11$ brightest "classical" dwarf
galaxies and has been confirmed as fainter satellites have been
detected. 
In the case of M31,  observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of the homogeneous
parent sample of $27$ satellites. 
However, taking into account the $11$ brightest satellites (as a
comparison with the MW) there isn't evidnce for any planar structure.

The question whether this kind of planar configurations are atypical
in the observable Universe cannot be properly addressed given the
current observational limitations.
Comparisons against observations in large galaxy surveys have only
been able to constrain that $5\%$ of the Milky-Way like galaxies in
the Universe have two bright satellite galaxies as is the case of the
Milky Way and its Large and Small Magellanic Clouds. 
A planar structure ouside the Local Group (LG) has only been confirmed
in Centaurus A \citep{muller2018whirling}.
